Stack Safety as a Security Property

### 2020-06-24 main.tex
Revamp intro and background

But what {\em is} stack safety? A formal definition of stack safety is
lacking from the literature, with the notion only being defined
negatively: through examples of stack unsafety. Two classical examples
of stack unsafety appear in Figure~\ref{fig:stackunsafety}.
\leo{Goal: on the left, a bad read; on the right, a bad write.}
\leo{Do we also need an example of ROP/control flow for the new part?}
\leo{Stack tokens introduce two properties: lse + wbcf, high level
  semantics, refinement to a machine with ``builtin'' ``by construction''
  (pretty weak :)) We make them formal and extensional}

A common undertone in such examples is the intention to protect
callers from callees: a callee should be not be able to access or
overwrite a caller's private data. In this paper we turn this
intuition into a formal security property, with a confidentiality and
an integrity component.
%
Informally, {\em stack confidentiality} ensures that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that if a function call returns, the caller's
stack data will be identical to what it was before the call.

@@

\leo{This will probably need to be fixed depending on what we end up
  doing}
%
A second common concern is to protect the call structure itself: a
callee should not be allowed to jump to the middle of an arbitrary
block of code either directly (via a jump instruction) or indirectly
(by corrupting stack data and returning). In this work we enforce
well-bracketed control flow by ensuring that code transitions are
valid based on a statically-known call graph of the
program. \leo{e.g. provided by the compiler?}

We begin by giving some background in Section~\ref{sec:prelim}.
Our contributions are:
\begin{itemize}
\item \leo{More? Extensional} formal characterizations of the intuitive concepts
  of local state encapsualtion \leo{ref} and well-bracketed control flow \leo{ref}.
  Evidence that this property is strong enough to capture the folk
  intuition: we show that the stack unsafe examples from the literature also fail
  our property.
\item Evidence that the property is not too strong. It's realizible by
  a previously proposed  micropolicy enforcement mechanism \leo{And it
    says yes on a substantial body of code}
  We then present a variation of the property + policy that is lazy,
  easier/more efficient to implement, and provably weaker.
\item A provably stronger (inductive? step-by-step/stepwise) variant
  of this property tuned for testing/verification.  We test the whole
  thing with a cool framework.
\end{itemize}
Finally, in Section~\ref{sec:relwork} we discuss related work and
sketch directions for future work in Section~\ref{sec:future}.

@@

We begin by describing our machine model. Concretely, we have in mind a
conventional machine architecture with a standard (RISC-V) processor
extended with some kind of (static or dynamic, software- or
hardware-enforced) security monitor, but we won't need to make any
specific choices about details until we get to \bcp{the policy one, probably}.
%In this section we will describe an abstract interface to our machine model.

Values and addresses are words drawn from some set of all words $\WORDS$. We will
denote words with $\word$.\leo{TODO: If we ever need ``A'' come back to this.}

Components are words (denoting addresses) and registers.
%
    \[\reg \in \REGS\]
    \[\component \in \COMPONENTS = \WORDS + \REGS \]
%
The registers must include two special purpose registers: the program
counter \PCname and the stack pointer \SP.

A machine state is just a map from components to values:
%
\[\mach \in \MACHS = \COMPONENTS \rightarrow \WORDS\]

Our stack safety property will be defined in terms of {\em observations}, which may be
silent (written $\tau$) or a word of data:
%
\[\obs \in \OBSS = \tau ~ | ~ \word \in \WORDS\]

These observations are a byproduct of the step function of the
machine, which takes a machine state and produces a pair of a
machine state and an observation.
%
\(\mach \stepstoobs{\obs} \mach' : \MACHS \rightarrow \MACHS \times \OBSS \).

\paragraph*{Policy}

A policy is a general model of an enforcement mechanism, consisting of
a set of policy states, \(\pol \in \POLS\) and a policy step function
\((\mach, \pol) \stepstopol \pol' : \MACHS \times \POLS
\rightharpoonup \POLS\).

%    The initializer and step function are both partial functions,
%    the former to reflect static guarantees (some initial states may not be valid at all) and
%    the latter to reflect dynamic fail-stop behavior. We then define machine-policy pairs and
%    a step function on them:

Let $\MPS$ be the set of all pairs of machine states and policy states.
%
\[\MPS = \MACHS \times \POLS\]
%
Then we can lift this policy step function to operate on such pairs,
by combining it with the regular step function for machine states.

%    \[\stepstoobs{}_P \subseteq \MPS \times \MPS\]

@@

\leo{Notation: Should $\stepstoobs{}$ be partial?}

@@

A trace over some type $A$ is a potentially infinite, nonempty
sequence of $A$s. We will mostly be concerned with sequences of
machines states, usually denoted $\machT$, and with sequences of
triples of (machine states, policy states, and observations), usually
denoted $\MPT$.

We define a few operations on traces generically.
The {\it split} operation \(\mathit{split}(f,T)\) takes a trace \(T\) and a predicate on elements \(f\) and
separates \(T\) into a prefix and suffix at the first element on which \(f\) holds:

  \judgment{\(f\ t\)}
           {\(\mathit{split}(f,tT) = t, tT\)}

  \judgmenttwo{\(\neg f\ t\)}{\(\mathit{split}(f,T) = T_1,T_2\)}
              {\(\mathit{split}(f,tT) = tT_1,T_2\)}

The {\it join} operation \(T_1 \cdot T_2\) takes two traces and links them by replacing the last element
of the first (if it has one) with the first of the second:

  \judgment{}
           {\(T_1t \cdot T_2 = T_1 T_2\)}

We define the ``trace-of'' operator \(\hookrightarrow\) that coinductively
relates a machine state with the trace of machine states paired with
observations produced by repeated application of step:
\[\machT \in \mathit{MOTrace} = \mathit{Trace}\ (\MACHS \times \OBSS)\]
\[\hookrightarrow : \machT \rightarrow \mathit{MOTrace}\]

\judgmenttwo{\(\mach_0 \stepstoobs{\obs} \mach_1\)}{\(\mach_1 \hookrightarrow \machT\)}
            {\(\mach_0 \hookrightarrow (\mach_0,\obs) \machT\)}

Since the step function is total, if \(\mach \hookrightarrow \machT\),
then \(\machT\) must be infinite.

\sna{Lambdas in the split are a bit busy, not sure anything will look better.}
We will also wish to take a prefix of the trace from an initial state up through the first state where
some condition holds, using {\it split}. Let \(f\) be a predicate on machine states. Then we define
\(\mach \hookrightarrow \machT | f\) (read ``\(\machT'\) is the prefix of \(\machT\) up to \(f\)''):

  \judgmenttwo{\(\mach \hookrightarrow \machT\)}
              {\(\mathit{split}(\lambda (\mach',\obs).f(\mach'),\machT) = \machT_1,\machT_2\)}
              {\(\mach \hookrightarrow \machT_1 | f\)}

Similarly, we define \(\hookrightarrow_P\) to relate a machine-policy
state pair to the trace of triples of machine states, policy states,
and observations induced by by \(\stepstoobs{}_P\):

\leo{I think this would read better as not exists. Thoughts?}
\sna{Looks good.}
  \[\MPT \in \mathit{MPOTrace} = \mathit{Trace}\ (\MACHS \times \POLS \times \OBSS)\]
  \[\hookrightarrow_P : \MACHS \times \POLS \rightarrow \mathit{MPOTrace}\]

  \judgment{\(\neg \exists \pol_1 . \pol_0 \rightharpoonup \pol_1\)}
           {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow \mpostate{\mach_0}{\pol_0}{\tau}\)}

  \judgmentthree{\(\mach_0 \stepstoobs{\obs} \mach_1\)}
                {\(\pol_0 \rightharpoonup \pol_1\)}
                {\(\mpstate{\mach_1}{\pol_1} \hookrightarrow \machT\)}
      {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow_P \mpostate{\mach_0}{\pol_0}{\obs} \machT\)}

We also lift the prefix-up-to relation to such traces.

@@

Finally, we will use \(\pi_m\) to project the trace of machine states
out of a trace, \(\pi_p\) to project the trace of policy states, and
\(\pi_o\) to project the observations.

@@

\paragraph*{Observations and Equivalence}

We say that a trace of observations $\obsT_1$ is a prefix of $\obsT_2$
\leo{Benjamin suggested ``as far as an attacker is conerned in an
  e-mail, but we haven't said anything about attackers yet.}, if the
sequence of word-observations of $\obsT_1$ is a prefix of those of
$\obsT_2$, that is we operate up to deletion of \(\tau\) observations, coinductively:

@@

Let the predicate \(\isinf(T)\) hold when the trace $T$ is infinite.
We then define {\em policy-sensitive observational equivalence}: an
{\it MPOTrace} \(\MPT\) is equivalent to an {\it MOTrace} trace \(\machT\) if \(\MPT\)
is infinite and the observations of \(\MPT\) and \(\machT\) coincide, or if the \(\MPT\)
is finite (due to a policy fault) and its observations are a prefix of those of M.

>>>>

But what {\em is} stack safety? A precise formal definition of stack
safety is lacking from the literature, with the notion only being
defined negatively: through examples of stack unsafety. Two classical
examples of stack unsafety appear in Figure~\ref{fig:stackunsafety}.
\leo{Goal: on the left, a bad read; on the right, a bad write.}
\leo{Do we also need an example of ROP/control flow for the new part?}
Recently, a \leo{inspiring/awesome/something good} work reasoned
about a stack safety enforcement mechanism using capabilities\leo{cite};
while impressive, stack safety was defined with respect to an abstract
machine that ``exhibits stack safety by construction''. Which begs
the question, what {\em is} the property that holds by construction?

\leo{StackTokens} provides some intuition towards answering that question,
by informally describing two components that comprise stack safety:
local state encapsulation and well-bracketed control flow. In this paper
we turn this intuition into two formal \leo{extensional} security properties,
that together form a formal definition of stack safety.

The underlying notion behind local state encapsulation is the
intention to protect callers from callees: a callee should be not be
able to access or overwrite a caller's private data. We formalize this
intuition as a security property, with a confidentiality and an
integrity component.
%
At a high level, {\em stack confidentiality} ensures that a callee's
observable behavior is not affected by memory outside its stack frame
(e.g. a caller's private data or uninitialized memory).  {\em Stack
  integrity} guarantees that if a function call returns, the caller's
stack data will be identical to what it was before the call.

@@

Similarly, the underlying notion behind well-bracketed control flow is
the need to protect the call structure itself: a callee should not be
allowed to jump to the middle of an arbitrary block of code either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). \leo{There needs to be some informal description
  of the wbcf part here. Sean?}

We begin by giving some background in Section~\ref{sec:prelim}.
Our contributions are:
\begin{itemize}
\item \leo{More? Extensional} formal characterizations of the
  intuitive concepts of local state encapsulation~\ref{sec:lse} and
  well-bracketed control flow~\ref{sec:wbcf}.  Evidence that this
  property is strong enough to capture the folk intuition: we show
  that the stack unsafe examples from the literature also fail our
  property.
\item Evidence that the property is not too strong. It's realizible by
  a previously proposed micropolicy enforcement
  mechanism~\ref{sec:micropolicy}. \leo{And it says yes on a
    substantial body of code} We then present a variation of the
  property + policy that is lazy, easier/more efficient to implement,
  and provably weaker.
\item A provably stronger (inductive? step-by-step/stepwise) variant
  of this property tuned for testing/verification.  We test the whole
  thing with a cool framework.
\end{itemize}
Finally, in Section~\ref{sec:relwork} we discuss related work and
sketch directions for future work in Section~\ref{sec:future}.

@@

Before we define stack safety, we need to first set the stage by
describing our machine model. Concretely, we have in mind a
conventional machine architecture with a standard (RISC-V) processor
extended with some kind of (static or dynamic, software- or
hardware-enforced) security monitor. However, our definition of stack
safety is generic: we will begin with an abstract interface to our
machine model and only make specific choices about details when we get
to the policy enforcement section~\ref{sec:micropolicy}.

\paragraph*{Values, States, and Observations}

The basic building blocks of the machine are {\em values} and {\em
  addresses}. Both are drawn from some set of all possible words
$\WORDS$. We will denote words with $\word$.
%
Our machine states are comprised of {\em components}, that can be word
(denoting memory addresses) or registers (drawn from some set $\REGS$.
%
    \[\reg \in \REGS\]
    \[\component \in \COMPONENTS = \WORDS + \REGS \]
%
The registers must include two special purpose registers: the program
counter \PCname and the stack pointer \SP.
%
Formally, a machine state is just a map from components to values:
%
\[\mach \in \MACHS = \COMPONENTS \rightarrow \WORDS\]

Our stack safety property will be defined in terms of {\em
  observations}, which may be silent (written $\tau$) or a word of
data:\leo{Needs more exposition}
%
\[\obs \in \OBSS = \tau ~ | ~ \word \in \WORDS\]
%
These observations are a byproduct of the step function of the
machine, which takes a machine state and produces a pair of a
machine state and an observation.
%
\(\mach \stepstoobs{\obs} \mach' : \MACHS \rightarrow \MACHS \times \OBSS \).

\paragraph*{Policies}

A policy is a general model of an enforcement mechanism, consisting of
a set of policy states \(\pol \in \POLS\), and a policy step function
\((\mach, \pol) \stepstopol \pol' : \MACHS \times \POLS
\rightharpoonup \POLS\).

We write $\MPS$ for the set of all pairs of machine states and policy
states.
%
\[\MPS = \MACHS \times \POLS\]
%
We then lift this policy step function to operate on such pairs,
by combining it with the regular step function for machine states.

@@

\leo{Notation: Should $\stepstoobs{}$ use harpoon for the lifted step?}

@@

A trace over some type $A$ is a potentially infinite, nonempty
sequence of $A$s. We will mostly be concerned with sequences of
machines states, usually denoted $\machT$, and with sequences of
triples of (machine states, policy states, and observations), usually
denoted $\MPT$. To continue, we will require three generic operations
on traces: a ``trace-of'' operation to construct the potentially
infinite traced induced by a step relation; a ``split'', to break a
trace into two components based on a predicate; a ``join'', the
reverse of split; and a ``prefix'', as a shorthand to only keep the
first component of a split.

The ``trace-of'' operator, written \(\hookrightarrow\), coinductively
relates a machine state with the trace of machine states paired with
observations produced by repeated application of step:
%
\[\hookrightarrow : \machT \rightarrow \MOTRACE\]%
%
\judgmenttwo{\(\mach_0 \stepstoobs{\obs} \mach_1\)}{\(\mach_1 \hookrightarrow \machT\)}
            {\(\mach_0 \hookrightarrow (\mach_0,\obs) \machT\)}%
%
\noindent
Since the step function is total, if \(\mach \hookrightarrow \machT\),
then \(\machT\) must be infinite.

Similarly, we define \(\hookrightarrow_P\) to relate a machine-policy
state pair to the trace of triples of machine states, policy states,
and observations induced by by \(\stepstoobs{}_P\):
%
\[\hookrightarrow_P : \MACHS \times \POLS \rightarrow \MPOTRACE \]
%
\judgment{\(\neg \exists \pol_1 . \pol_0 \rightharpoonup \pol_1\)}
         {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow \mpostate{\mach_0}{\pol_0}{\tau}\)}
%  
\judgmentthree{\(\mach_0 \stepstoobs{\obs} \mach_1\)}
              {\(\pol_0 \rightharpoonup \pol_1\)}
              {\(\mpstate{\mach_1}{\pol_1} \hookrightarrow \machT\)}
              {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow_P \mpostate{\mach_0}{\pol_0}{\obs} \machT\)}
We will also use \(\pi_m\) to project the trace of machine states
out of a trace, \(\pi_p\) to project the trace of policy states, and
\(\pi_o\) to project the observations.

The {\it split} operation \(\mathit{split}(f,T)\) takes a trace \(T\)
and a predicate on elements \(f\) and separates \(T\) into a prefix
and suffix at the first element on which \(f\) holds:

\judgment{\(f\ t\)}
         {\(\mathit{split}(f,tT) = t, tT\)}
%  
\judgmenttwo{\(\neg f\ t\)}{\(\mathit{split}(f,T) = T_1,T_2\)}
            {\(\mathit{split}(f,tT) = tT_1,T_2\)}
%
Note that the last element of the prefix and the first element of the
suffix coincide. As we will mostly be using $\mathit{split}$ to
partition a trace to a callee and a caller portion, this choice will
allow us to preserve local state encapsulation invariants during
the transition between the two.

The {\it join} operation, written \(T_1 \cdot T_2\) is the reverse of
$\mathit{split}$: it takes two traces and links them by replacing the
last element of the first (if it has one) with the first element of
the second

\judgment{} {\(t \cdot T = T\)}
\judgment{\(T_1 \cdot T_2 = T\)}{\(tT_1 \cdot T_2 = tT\)}
\leo{Lemma/theorem: split/join = id?}
\leo{TODO: FORMATTING: Remove force line break/par at end of judgments}

Finally, we will also want to take a prefix of the trace from an
initial state up through the first state where some condition holds.
This can easily be implemented using $\mathit{split}$, but as we're
going to be making heavy use of it we're going to provide special
notation for convenience.
%
Let \(f\) be a predicate on machine states. Then we define \(\mach
\hookrightarrow \machT | f\) (read ``\(\machT'\) is the prefix of
\(\machT\) up to \(f\)''):

\judgmenttwo{\(\mach \hookrightarrow \machT\)}
            {\(\mathit{split}(\lambda (\mach',\obs).f(\mach'),\machT)
              = (\machT_1,\machT_2)\)}
            {\(\mach \hookrightarrow \machT_1 | f\)}
\noindent
And similarly for traces with policies:

@@

@@

\paragraph*{Observational Equivalence}

We say that a trace of observations $\obsT_1$ is a prefix of $\obsT_2$
\leo{Benjamin suggested ``as far as an attacker is conerned in an
  e-mail, but we haven't said anything about attackers yet.}, if the
sequence of non-silent observations of $\obsT_1$ is a prefix of those
of $\obsT_2$; that is we operate up to deletion of \(\tau\)
observations, coinductively:

@@

Let the predicate \(\isinf(T)\) hold when the trace $T$ is infinite.
We then define {\em policy-sensitive observational equivalence}: a
trace $\MPT \in \MPOTRACE$ is equivalent to a trace $\machT \in
\MOTRACE$ if (1) \(\MPT\) is infinite and the observations of \(\MPT\)
and \(\machT\) are equivalent, or (2) if the \(\MPT\) is finite (due to a
policy fault) and its observations are a prefix of those of M.
